A Flux Scale for Southern Hemisphere 21cm EoR Experiments

### 2013-04-23 ms.tex
Minor edits (except for substantial stuff in abstract), commented in areas to focus on and proposed changes.  Will proceed shortly with trying to execute these various changes.

\shorttitle{PSA64}
\shortauthors{Jacobs et al.}

@@

%\title{The Reionization flux scale in the Southern Sky}
\title{A southern hemisphere flux scale for 21cm epoch of reionization experiments}
\author{
Daniel C. Jacobs\altaffilmark{1},
Aaron R. Parsons\altaffilmark{2},
James E. Aguirre\altaffilmark{3},
Zaki Ali\altaffilmark{2},
Richard F. Bradley\altaffilmark{4,5,6},
Chris L.  Carilli\altaffilmark{7},
Dave DeBoer\altaffilmark{8},
Matthew Dexter\altaffilmark{8},
Nicole E. Gugliucci\altaffilmark{5},
Dave MacMahon\altaffilmark{8},
Jason Manley\altaffilmark{9},
David F. Moore\altaffilmark{3},
Jonathan C. Pober\altaffilmark{2},
Irina I. Stefan\altaffilmark{10},
William P. Walbrugh\altaffilmark{9}}

@@

\begin{abstract}
We present the first spectral measurements from the Precision Array for Probing
the Epoch of Reionization (PAPER) in South Africa observed in July and
September of 2011.  The spectra are determined by calibrating directly to a
nearby source, eliminating the error-prone primary beam correction.  The global
flux calibration, bootstrapped from catalog data of 5 bright sources, is
estimated to be accurate to 3.2\%.  Comparing with prior measurements, 45 out
of 58 sources observed are found to confirm and refine a spectral index model.
This includes Pictor A which provides a key flux reference for PAPER's EoR
power spectrum. Combining catalog data at other frequencies with the PAPER data
we are able to estimate the flux model parameters to an uncertainty of 2.4\%.
This accuracy, which is ultimately limited by the uncertainty of the many prior
measurements used to estimate the flux scale, is an order of magnitude
improvement on previous estimates.  
\end{abstract}

@@

Numerous radio telescopes are now exploring the prospects for using
measurements of highly redshifted 21cm emission to inform our understanding of
cosmic reionization in the redshift range $6< z<12$, corresponding to radio
frequencies below 200MHz (see reviews in
\citet{Furlanetto:2006p2267,Morales:2010p8093,Pritchard:2012p9555}) .  These
include telescopes aiming to measure the global temperature change of 21cm
emission during the Epoch of Reionization (EoR), such as the Compact
Reionization Experiment (CoRE) and the Experiment to Detect the Global EoR
Signature (EDGES; \citealt{Bowman:2010p8546}), and interferometers aiming to
measure the power spectrum of 21-cm EoR emission, such as the Giant Metre-wave
Radio Telescope (GMRT;
\citealt{Paciga:2011p9470,Paciga:2013p9627})\footnote{\url{http://gmrt.ncra.tifr.res.in/}},
the LOw Frequency ARray (LOFAR;
\citealt{Yatawatta:2013p9699})\footnote{\url{http://www.lofar.org/}}, the
Murchison Widefield Array (MWA;  \citealt{Bowman:2012p9138};
\citealt{Tingay:2013p9022}))\footnote{\url{http://www.mwatelescope.org/}}, and
the Donald C. Backer Precision Array for Probing the Epoch of Reionization
(PAPER; \citealt{Parsons:2010p6757})\footnote{\url{http://eor.berkeley.edu/}}.
Both PAPER and the MWA are located in the southern hemisphere at radio quiet
reserves being prepared for the upcoming Square Kilometer Array.  At these low
frequencies the Southern sky is much less well known than the North, with the
most extensive surveying work being conducted  by the EoR experiments
themselves \citep{Jacobs:2011p8438,Williams:2012p8768}.

Neutral hydrogen is visible during reionization as $\sim$30mK excess surface
brightness over the CMB punctuated both spectrally and spatially by regions of
zero emission.  This patchy signal is predicted to be visible everywhere on the
sky at size scales up to degrees, or about $0.1$cMpc$^{-1}$.   To increase the
surface brightness sensitivity without building costly dishes, MWA, PAPER and
LOFAR have chosen to use combinations of phased and/or correlated dipoles. By
design also cover much wider fields of view ($>10$\arcdeg) and bandwidths
($\sim 100$\% fractional) than traditional dish telescopes.   As the elements
can not be physically pointed, the apparent flux scale varies across the sky.
This direction dependent gain is currently uncertain to 20\% or higher
\citep{Jacobs:2013p9713}.  The ideal solution is to use a wide-grid of
well-known calibrators to map the primary beam and set the flux scale.

@@

However, catalog source fluxes at 150MHz are also inaccurate to the 20\% level,
particularly at southern declinations ($<-20$\arcdeg) . Historically, the best
EoR band data were by \citet{Slee:1995p7541} with Culgoora Circular
Array\footnote{Known during daylight hours as Culgoora Radio Heliograph} and
various higher frequency measurements with Parkes.  These data are typically
uncertain to 10\% or higher and provide little coverage of the EoR band beyond
a single narrow-band data point.

PAPER, in its maximum redundancy configuration, has a very broad beam with
large regular grating side-lobes and is therefor limited to using the few
sources above $\sim$50 Jy for flux calibration. The best calibrator for most
observations is Pictor A. For this source the current best EoR band measurement
is uncertain to 12\% and appears to imply spectral flattening in the EoR band
\citep{Perley:1997p9312}.

@@

The spectra were measured by PAPER, a dedicated EoR experiment 
that employs drift-scanning, dual-polarization dipole antennas 
tuned for efficient operation over a 120--170-MHz band located in the South African Karoo desert
on the Square Kilometer Array
South Africa (SKA-SA) reserve, 100km north of the small town of Carnarvon.
The PAPER array has grown from 16 elements deployed in early 2009 to a
64-element imaging array in 2011 (see Figure \ref{fig:antpos}). Since November 2011 it has been arranged in a maximally redundant grid configuration to make deep power spectral integrations.  
\begin{figure*}\centering
\includegraphics[width=0.85\columnwidth]{plots/antpos.png}
\caption{The 64-antenna, minimum-redundancy PAPER array at Karoo Radio Observatory site in South Africa used for these observations.
}\label{fig:antpos}
\end{figure*}

In this paper our primary goal is to provide a subset of accurate fluxes
suitable for EoR power spectrum and catalog calibration.  After recording the
data used in this paper, PAPER was reconfigured into a redundant layout
optimized to detect a small number of power spectrum bins at high precision, an
arrangement which trades imaging dynamic range for power spectrum sensitivity.
As such it is only able to easily distinguish the very brightest sources on the
sky a step necessary for setting the overall flux scale. Of these, Pictor A is
one of the most useful, being bright, minimally resolved and far from from
confusing structure.

@@

@@

\begin{figure}
\includegraphics[width=0.96\textwidth]{plots/picA_Perley.png}
\caption{
Pictor A imaged at 330MHz by the VLA \citep{Perley:1997p9312}. Black circle
indicates PAPER resolution at the 150MHz, dashed, the resolution at 185MHz.  To
correct for the small residual structure in the PAPER measurements we
interpolated the PAPER baselines in the uv plane. At the highest frequency this
led to a 3\% correction, at the lowest frequency, 0.6\%.
\label{fig:pic_perley}}
\end{figure}

@@

The primary purpose of this observation is to produce a high quality Pictor A
spectrum. As a verification measure we have observed many other sources at the
same declination for comparison against previous observations.  Our comparison
metric will be the degree to which the PAPER data are consistent with previous
observations,given a model and, if consistent, they improve on the precision of
our spectral model parameters.

@@

Inspecting the confidence contours we find that, for the vast majority ($\sim
75$\%), the PAPER measurements confirm the power law extrapolation from higher
frequencies. The remaining 25\% either A) have large PAPER error bars and
therefor provide no new information or B) do not agree with the spectral index
model.  To understand where most measurements fall on this spectrum we
numerically quantify the overall model improvement derived from the addition of
the PAPER data as 
\begin{align}
\textrm{Improvement} &= \left(\textrm{Precision increase}\right) \left(\textrm{Catalog agreement}\right)\\
&= \left(\frac{1}{Area(\textrm{PAPER})} - \frac{1}{Area(\textrm{Cat})}\right) \left( Area(\PAPER \cap \textrm{Cat}) \right)
\end{align}

@@

However, the majority of source 78\% have positive improvement index,
indicating strong confirmation of the extrapolated spectrum (see Figures
\ref{fig:SI_contour_1} - \ref{fig:SI_contour_4}).  Pictor A is in the middle of
this group  with an improvement of 0.39 ---only 14 sources show stronger
confirmation.  The flux model is $S150=$388 $\pm$ 9.4Jy, $\alpha=-0.77\pm0.02$.
The combination of several independent PAPER data points near 150MHz with
previous catalog measurements results in this estimated uncertainty of 2.4\%.
This uncertainty is dominated by agreement between the 2 Wills measurements
near 600MHz which are accurate to 2\% and the five PAPER measurements, which
are accurate to ~5\%.

@@

@@

\section*{Appendix: Model Fitting}

>>>>

\shorttitle{Southern Hemisphere Flux Scale}
\shortauthors{Jacobs et al.}

@@

%\title{The Reionization flux scale in the Southern Sky}
\title{A Flux Scale for Southern Hemisphere 21cm EoR Experiments}
\author{
Daniel C. Jacobs\altaffilmark{1},
Aaron R. Parsons\altaffilmark{2},
James E. Aguirre\altaffilmark{3},
Zaki Ali\altaffilmark{2},
Richard F. Bradley\altaffilmark{4,5,6},
Chris L.  Carilli\altaffilmark{7},
David R. DeBoer\altaffilmark{8},
Matthew R. Dexter\altaffilmark{8},
Nicole E. Gugliucci\altaffilmark{5},
Pat Klima\altaffilmark{5},
Dave H. E. MacMahon\altaffilmark{8},
Jason R. Manley\altaffilmark{9},
David F. Moore\altaffilmark{3},
Jonathan C. Pober\altaffilmark{2},
Irina I. Stefan\altaffilmark{10},
William P. Walbrugh\altaffilmark{9}}

@@

\begin{abstract}
We present a first catalog of broad-band spectral measurements from the 
Donald C. Backer Precision Array for Probing
% XXX ARP: was irina "first"?  and maybe "catalog"
the Epoch of Reionization (PAPER) in South Africa observed in July and
September of 2011.  In order to reduce the impact of beam calibration, which proves to
be a difficult endeavor for transit telescopes such as PAPER, on the determination
of source spectra, we have focused on calibrating sources in a narrow declination range.
Since each source follows a nearly identical path through the primary beam, this
restriction allows beam calibration to be nearly eliminated as a source of error,
yielding a dramatic improvement in the accuracy of source spectra measured in the 
100--200-MHz band that is receiving renewed attention by experiments seeking to
measure 21cm emission from the Epoch of Reionization (EoR).
Based on the variation in our data and the estimated error in the absolute
flux scale we bootstrap from multiple bright sources, we estimate PAPER measurements
of sources above XXX Jy to be accurate to 3.2\%.
When combined with catalog data at other frequencies, these data constrain parameters of
a power-law model for flux density with an uncertainty of 2.4\%.
% XXX ARP: this must again be dependent on source strength and location.
This accuracy is limited by the uncertainty in the catalog measurements we use
to estimate an absolute flux scale, and represents 
an order of magnitude improvement over previous measurements in this band.  
Comparing with prior measurements, 45 out
of 58 sources observed are found to confirm and refine a power-law model for flux density.
This includes Pictor A, which provides a key flux reference for PAPER's EoR
power spectrum analysis.
\end{abstract}

@@

Numerous radio telescopes are now exploring the prospects for using
measurements of highly redshifted 21cm emission to inform our understanding of
cosmic reionization in the redshift range $6< z<12$, corresponding to radio
frequencies below 200 MHz (see reviews in
\citealt{Furlanetto:2006p2267,Morales:2010p8093,Pritchard:2012p9555}) .  These
include telescopes aiming to measure the global temperature change of 21cm
emission during the Epoch of Reionization (EoR), such as the Compact
Reionization Experiment (CoRE) and the Experiment to Detect the Global EoR
Signature (EDGES; \citealt{Bowman:2010p8546}), and interferometers aiming to
measure the power spectrum of 21-cm EoR emission, such as the Giant Metre-wave
Radio Telescope (GMRT;
\citealt{Paciga:2011p9470,Paciga:2013p9627})\footnote{\url{http://gmrt.ncra.tifr.res.in/}},
the LOw Frequency ARray (LOFAR;
\citealt{Yatawatta:2013p9699})\footnote{\url{http://www.lofar.org/}}, the
Murchison Widefield Array (MWA;  \citealt{Bowman:2012p9138};
\citealt{Tingay:2013p9022}))\footnote{\url{http://www.mwatelescope.org/}}, and
the Donald C. Backer Precision Array for Probing the Epoch of Reionization
(PAPER; \citealt{Parsons:2010p6757,Parsons2013b})\footnote{\url{http://eor.berkeley.edu/}}.
Both PAPER and the MWA are located in the southern hemisphere at radio quiet
reserves being prepared for the upcoming Square Kilometer Array.  At these low
frequencies the Southern sky is much less well known than the North, with the
most extensive surveying work being conducted  by the EoR experiments
themselves \citep{Jacobs:2011p8438,Williams:2012p8768}.

Neutral hydrogen is visible during reionization as $\sim$30mK excess surface
brightness over the CMB punctuated both spectrally and spatially by regions of
zero emission.  This patchy signal is predicted to be visible everywhere on the
% XXX ARP: don't like description of EoR signal as "surface brightness"
sky at size scales up to degrees, or about $0.1$cMpc$^{-1}$.   To increase the
surface brightness sensitivity without building costly dishes, MWA, PAPER and
LOFAR have chosen to use combinations of phased and/or correlated dipoles. By
design also cover much wider fields of view ($>10$\arcdeg) and bandwidths
($\sim 100$\% fractional) than traditional dish telescopes.   As the elements
% XXX what about GMRT?
can not be physically pointed, the apparent flux scale varies across the sky.
This direction dependent gain is currently uncertain to 20\% or higher
\citep{Jacobs:2013p9713}.  The ideal solution is to use a wide-grid of
well-known calibrators to map the primary beam and set the flux scale.  
% XXX don't go to "ideal solution" just yet
% XXX ARP: add importance of accurate flux vs. frequency (Parsons et al. 2012, etc.)

@@

However, catalog source fluxes at 150MHz are also inaccurate to the 20\% level,
particularly at southern declinations ($\delta<-20$\arcdeg) . Historically, the best
EoR band data were by \citet{Slee:1995p7541} with Culgoora Circular
Array\footnote{Known during daylight hours as Culgoora Radio Heliograph} and
various higher frequency measurements with Parkes.  These data are typically
uncertain to 10\% or higher and provide little coverage of the EoR band beyond
a single narrow-band data point.

PAPER, in its maximum redundancy configuration, has a very broad beam with
% XXX need background on minimum/maximum array configurations
large regular grating side-lobes and is therefore limited to using the few
sources above $\sim$50 Jy for flux calibration. The best calibrator for most
% XXX qualify "best"
observations is Pictor A. For this source the current best EoR band measurement
is uncertain to 12\% and appears to imply spectral flattening in the EoR band
\citep{Perley:1997p9312}.

@@

The spectra were measured by PAPER, a dedicated EoR experiment 
that employs drift-scanning, dual-polarization dipole antennas 
tuned for efficient operation over a 120--170-MHz band located in the South African Karoo desert
on the Square Kilometer Array
South Africa (SKA-SA) reserve, 100km north of the small town of Carnarvon.
The PAPER array has grown from 16 elements deployed in early 2009 to a
64-element imaging array in 2011 (see Figure \ref{fig:antpos}). Since November 2011 it has been arranged in a maximally redundant grid configuration to make deep power spectral integrations \citep{Parsons:2012a}.
\begin{figure*}\centering
\includegraphics[width=0.85\columnwidth]{plots/antpos.png}
\caption{Antenna position in the 64-antenna, minimum-redundancy PAPER array configuration at the Karoo Radio Observatory site in South Africa used for these observations.
% XXX add brief summary of observations.
}\label{fig:antpos}
\end{figure*}

In this paper our primary goal is to provide a subset of accurate fluxes
suitable for EoR power spectrum and catalog calibration.  After recording the
data used in this paper, PAPER was reconfigured into a redundant layout
optimized to detect a small number of power spectrum bins at high precision, an
arrangement which trades imaging dynamic range for power spectrum sensitivity.
As such it is only able to easily distinguish the very brightest sources on the
sky a step necessary for setting the overall flux scale \citep{Parsons2013b}. Of these, Pictor A is
one of the most useful, being bright, minimally resolved and far from from
confusing structure.

@@

\section{Approach}

% XXX write this section

@@

\begin{figure}
\includegraphics[width=0.96\textwidth]{plots/picA_Perley.png}
\caption{
Pictor A imaged at 330MHz by the VLA \citep{Perley:1997p9312}. Black circle
indicates PAPER resolution at the 150MHz, dashed, the resolution at 185MHz.  To
correct for the small residual structure in the PAPER measurements we
interpolated the PAPER baselines in the uv plane. At the highest frequency this
% XXX "interpolated" needs to be clarified
led to a 3\% correction, at the lowest frequency, 0.6\%.
\label{fig:pic_perley}}
\end{figure}

@@

The primary purpose of this observation is to produce a high quality Pictor A
% XXX need to coordinate "primary purpose" here with abstract and intro motivation
spectrum. As a verification measure we have observed many other sources at the
same declination for comparison against previous observations.  Our comparison
metric will be the degree to which the PAPER data are consistent with previous
observations,given a model and, if consistent, they improve on the precision of
our spectral model parameters.

@@

Inspecting the confidence contours we find that, for the vast majority ($\sim
75$\%), the PAPER measurements confirm the power law extrapolation from higher
frequencies. The remaining 25\% either A) have large PAPER error bars and
therefore provide no new information or B) do not agree with the spectral index
model.  To understand where most measurements fall on this spectrum we
numerically quantify the overall model improvement derived from the addition of
the PAPER data as 
\begin{align}
\textrm{Improvement} &= \left(\textrm{Precision increase}\right) \left(\textrm{Catalog agreement}\right)\\
&= \left(\frac{1}{Area(\textrm{PAPER})} - \frac{1}{Area(\textrm{Cat})}\right) \left( Area(\PAPER \cap \textrm{Cat}) \right)
\end{align}
% XXX This equation is yucky

@@

However, the majority of source 78\% have positive improvement index,
indicating strong confirmation of the extrapolated spectrum (see Figures
\ref{fig:SI_contour_1} - \ref{fig:SI_contour_4}).  Pictor A is in the middle of
this group  with an improvement of 0.39 ---only 14 sources show stronger
confirmation.  The flux model is $S150=$388 $\pm$ 9.4Jy, $\alpha=-0.77\pm0.02$.
% XXX need more detail here.  This is the crux of what is done.
The combination of several independent PAPER data points near 150MHz with
previous catalog measurements results in this estimated uncertainty of 2.4\%.
This uncertainty is dominated by agreement between the 2 Wills measurements
near 600MHz which are accurate to 2\% and the five PAPER measurements, which
are accurate to ~5\%.

@@

% XXX need more discussion

@@

\section*{Appendix: Model Fitting}
% XXX this should be in main body, not an appendix
